BART: A Comparison with Frequentist Tree Methods

### 2025-01-10 paper/paper.tex
updated draft

\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{listing}
\usepackage{color}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}
\usepackage{booktabs}
\usepackage{caption}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{subcaption}
\usepackage{bookmark}
\usepackage[
    backend=biber,
    style=authoryear,
  ]{biblatex}

@@

\title{BART: A Comparison with Non-bayesian Tree Methods}
\author{
    Vincenzo Dorrello 
    \and
    Giulio Frey 
    \and
    Guido Rossetti 
    \and
    Giovanni Scarpato 
}
\date{\today}

@@

\begin{abstract}
This paper provides a comprehensive comparison between Bayesian Additive Regression Trees (BART) and traditional non-Bayesian tree-based methods for regression problems. We begin by examining the theoretical foundations of decision trees and various ensemble methods, including bagging, random forests, and boosting. We then present BART as a Bayesian nonparametric approach that combines the flexibility of regression trees with the formal uncertainty quantification of Bayesian inference. The paper details BART's probability model, its regularization prior, and the Bayesian backfitting MCMC algorithm used for posterior inference. Through both simulated and real-world data applications, we demonstrate BART's effectiveness in capturing non-linear relationships and compare its predictive performance against other ensemble methods. Using the UCI Abalone dataset, we show that BART achieves superior prediction accuracy compared to random forests, boosting, and bagging when using sufficient numbers of trees, though at a higher computational cost. Our findings suggest that BART's automatic prior-based regularization and ability to quantify uncertainty make it a valuable addition to the tree-based regression toolkit, particularly for complex non-linear modeling tasks.
\end{abstract}

@@

The challenge of modeling complex, non-linear relationships in data has led to the development of various tree-based methods in statistical learning. While traditional parametric approaches like linear regression impose strict assumptions on the relationship between predictors and outcomes, tree-based methods offer greater flexibility in capturing non-linear patterns and interaction effects. This paper focuses on comparing Bayesian Additive Regression Trees (BART) with traditional non-Bayesian tree methods, examining their theoretical foundations, implementation approaches, and practical performance.

Tree-based methods have evolved from simple decision trees to sophisticated ensemble approaches such as bagging, random forests, and boosting. These methods combine multiple trees to improve prediction accuracy and reduce overfitting. However, most traditional approaches lack formal uncertainty quantification and require careful tuning of regularization parameters. BART, introduced by \cite{chipmanBARTBayesianAdditive2010}, addresses these limitations by providing a Bayesian framework that automatically handles regularization through prior specifications while offering natural uncertainty estimates.

@@

The paper is organized as follows: Section \ref{decision} introduces decision trees and their fundamental concepts. Section \ref{frequentist} explores frequentist ensemble methods including bagging, random forests, and boosting. Section \ref{bart} provides a detailed examination of BART, including its model specification, prior structure, and posterior sampling approach. Section \ref{Comparison} offers a theoretical comparison of these methods, highlighting their relative strengths and weaknesses. Finally, Section \ref{data_app} presents empirical results from both simulated and real-world data applications, demonstrating the practical implications of choosing between these methods for regression tasks.\footnote{The code repository for this paper is available at this \href{https://github.com/giuliofrey/bart}{link}}

@@

Consider a target variable \( Y \) and a \( p \)-dimensional vector of covariates, \( x = (x_1, \ldots, x_p) \). Decision trees are predictive models that make predictions about \( Y \) by partitioning the predictor space—the set of all possible values of \( x \)—into simple, non-overlapping regions. Notice that also linear
regression in a sense partitions the predictor space, but restricts divisions to certain planes. To the contrary, all trees requires is that the partition can be achieved by successive binary splits based on the different predictors. Once we have a partition such as this we base our prediction on the average of the $Y$ in each partition.

The structure of a decision tree reflects this partitioning process. The tree begins with a root node at the top, representing the entire dataset (or the full covariate space). The root node is split into two subsets according to decision rules based on the predictor variables \( (x_1, \ldots, x_p) \). Each subsequent subset may also be further split into two new subsets, continuing the process recursively. At each split, a new internal node is created, and the segments connecting nodes are called branches. This recursive splitting process, known as \textit{recursive partitioning}, proceeds iteratively until a termination condition is met, such as when further splitting no longer improves predictions or the subsets become too small. The final nodes, called \textit{terminal nodes} or \textit{leaves}, correspond to the smallest subsets created by the tree.

The terminal nodes define a partition of the predictor space into \( J \) non-overlapping regions, \( R_1, R_2, \ldots, R_J \). Each region corresponds to a leaf of the tree. To make a prediction for a given covariate vector \( x \), the tree is traversed starting at the root node. At each internal node, the splitting rule determines whether to proceed to the left or right child node. This process continues until a terminal node is reached. The predicted value \( \hat{y} \) is then assigned based on the region \( x \) belongs to.

When \( Y \) is continuous, the tree is called a \textit{regression tree}, and the predicted value \( \hat{y}_i \) for a new observation \( x_i \) falling into region \( R_j \) is the mean of \( Y \) values for the training observations in \( R_j \). For categorical \( Y \), the tree is referred to as a \textit{classification tree}, and the predicted value is the majority class of \( Y \) among the training observations in \( R_j \). Collectively, these approaches are known as \textit{CART} (Classification and Regression Trees), a term introduced by \cite{Gordon_Breiman_Friedman_Olshen_Stone_1984}.

In this work, we focus specifically on regression trees, which aim to predict a continuous \( Y \) given a training dataset. Regression trees differ fundamentally from linear regression models because they do not impose any parametric form on the relationship between \( x \) and \( Y \). Instead, regression trees use a piecewise constant approximation of this relationship.

To see this, suppose there are \( n \) observations of \( Y \) and \(\mathbf{x} = (x_1, \ldots, x_p) \). The relationship between the covariates and the response variable can be expressed as:
\begin{equation}
Y = f(X) + \epsilon \label{regression}
\end{equation}
where \( \epsilon \) is an error, whose distribution is not necessarily specified. Both regression trees and linear regression models seek to estimate the unknown function \( f(X) \) using the observed data.

The manner in which \( f(X) \) is approximated, however, depends on the model chosen. Linear regression assumes a parametric form:
\[
f(X) = \beta_0 + \sum_{j=1}^p X_j \beta_j,
\]
where \( \beta_0, \beta_1, \ldots, \beta_p \) are coefficients estimated from the data. In contrast, regression trees approximate \( f(X) \) as a stepwise constant function:
\[
f(X) = \sum_{j=1}^J \mu_j \cdot 1(X \in R_j),
\]
where \( R_1, \ldots, R_J \) are the regions defined by the leaves of the tree, and \( \mu_j \) is the mean of \( Y \) for the training observations in region \( R_j \).

The choice between these models depends on the nature of the underlying relationship between the predictor variables and the response variable. If the relationship is approximately linear, a linear regression model is likely to outperform a regression tree. Conversely, if the relationship is highly nonlinear or complex, regression trees can capture such patterns more effectively. Additionally, regression trees have the advantage of being invariant to the scale of the predictor variables, eliminating the need for preprocessing steps like standardization.

This section draws from \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021} and \cite{Genuer_Poggi_2020} to illustrate these concepts.

\subsection{Regression Trees}
A regression tree partitions the \(X\)-space into disjoint regions \(R_j\) and provides a fitted value \(E(Y \mid X \in R_j)\) within each region.
We consider the following notation. Denote by \( T \) the tree structure, consisting of a set of interior node decision rules and a set of $b$ terminal nodes. The decision rules are binary splits of the predictor space of the form \( \{x \in A\} \) vs \( \{x \notin A\} \), where \( A \) is a subset of the range of \( x \). We only consider rules based on the single components of \( x = (x_1, \dots, x_p) \) and of the form \( \{x_i \leq c\} \) vs \( \{x_i > c\} \) for continuous \( x_i \). Let \( M = \{\mu_1, \mu_2, \dots, \mu_b\} \) denote a set of parameter values associated with each of the \( b \) terminal nodes of \( T \). Given the way it is constructed, the tree is a full binary tree, that is, each node has exactly zero or two children.

We now delve better in the process of building regression trees. As we saw before, they are built in two steps. First, the predictor space (the set of all possible values of the vector of observables $X_1, ...., X_p$) is split into $J$ regions $R_j$ that are distinct and non-overlapping. Secondly, given a new observation $x$ that falls in a region $R_j$, the model predicts the corresponding outcome variable $\hat{y}$ to be the mean of the outcome variables corresponding to the observations in the training set that fall in that same region.

To begin, we have to construct the regions $R_1,..., R_j$. Typically, regions are chosen as $p$-dimensional boxes, where $p$ is the number of regressors. The partition is chosen is such a way to minimize the RSS:  
\begin{equation}
  \{R_1,...,R_J \}=\operatorname*{argmin}_{ \{R_j\}_{j=1}^J}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\mu_{R_j}\right)^2
  \label{rss_tree}
\end{equation}
where $\mu_{R_j}$ is the mean of the training observation in the $j$-th box.

It is computationally infeasible to do the optimization problem in \eqref{rss_tree}. Instead of considering all the possible partitions of the predictor space, a top-down, greedy approach, known as recursive binary splitting is used. The approach is \texttt{top-down} since it begins at the top of the tree, and then sequentially splits the predictor space; each split is then indicated by two new branches (hence it is binary). Moreover, the process is \textit{greedy}, since for each step we make the split that generates the greatest reduction of the RSS at that particular step, and we do not choose the best sequence of steps looking ahead.
\\To perform recursive binary splitting, the algorithm considers the set of all predictors $X_1, \ldots, X_p$, and the set of all possible values of the cutpoint $s$, and then choose the predictor $X_j$ and the cutpoint $s$ such that the branching of the tree corresponding to the decision rule:  $$\{X \mid X_j < s\} \quad \text{and} \quad \{X \mid X_j \geq s\}$$ minimizes the RSS. Then, we repeat the process within the resulting subregion, looking again for the predictor and the cutpoint such that the subsequent decision rule minimize the RSS. It is important to note that this time we do not split the entire predictor space but one of the two previously identified regions. This process continues until a stopping condition is reached; the more common requires a minimum number of observation within each splitting nodes.

\subsection{Fitting trees: growing and pruning}

Recursive binary splitting to build a regression tree  is likely to produce good prediction on the training-set but overfit the training data, leading to poor out-of-sample performance.The reason is that the above mentioned procedure is likely to give us a complex tree, which has probably a low bias but high variance in prediction. Indeed, small fluctuations in the training set may lead to different predictions.

Constraining ex ante the size of trees would be a short-sighted solution, because one could possibly stop too soon, leaving potentially good splits further down in tree unachieved.
\\A better strategy is pruning an already complex tree. This amounts to search for a good intermediate between two extreme cases: a fully developed tree which has high variance but low bias, and a tree consisting of only the root, with high bias but low variance. 
The strategy consists in growing a very large tree $T_0$, and then prune it back in order to obtain a subtree. To do so, we use cost complexity pruning, also know as weakest link pruning.

Consider a sequence of subtrees of $T_0$, $\{T\}_\alpha$ indexed by a non-negative tuning parameter $\alpha$.
For each value of $\alpha$, there is a subtree $T \subset T_0$ such that
$$\sum_{m=1}^{|T|} \sum_{i : x_i \in R_m} \left( y_i - \mu_{R_m} \right)^2 + \alpha |T|$$
is as small as possible, where $|T|$ represents the number of terminal nodes of the tree T, $R_m$ is the subset of the predictor space corresponding to the $m$-th terminal node, $\mu_{R_m}$ is the predicted response associated with $R_m$.
The tuning parameter $\alpha$ controls the trade-off between a sub-tree complexity and and its fit to the training data, the higher the $\alpha$, the more terminal nodes a tree has, the smaller a subtree is.
The value for $\alpha$ can be chosen using a validation set or trough cross-validation.

\subsection{Pros and Cons of decision trees}
Overall, regression trees offer much greater flexibility in handling the relation between covariates and output than linear models do. This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns. They can more naturally incorpo
rate interaction effects.

Greater flexibility comes in handy also when dealing with missing values. Tree offer an effective way to overcome this issue, by constructing surrogate variables. Exploiting correlation between predictors, they may decrease the effect of missing data.

So far we considered only binary splits, thus an interesting question would be if a model with more than two splits at each note is a valid idea. However, it turns out that in general is not a good strategy. The issue is that multiple splits fragment the dataset too quickly. Moreover, since multiple splits can be achieved by a series of binary splits, the latter approach is preferred.

The interpretability of trees has been one of the driving forces of its success. Indeed, thanks to their structure, is straightforward to understand why, for a given $x$, we expect a particular value of $y$, following a series of decision rules. Small trees are visually easy to interpret. Moreover, trees build in such a way are a computationally fast method, that easily scales with large datasets.

However, one of the main disadvantages of this setting, is that trees are non-robust. They suffer from high variance: a small change in the data can lead to a large change in the estimation. The main reason of this instability is the hierarchical structure of the process as the effect of an error in one of the first nodes is propagated to the other nodes below it. This is heightened by the fact that single trees tend to overfit, and spurious correlations in the training data can be mistaken as structural patterns and included in the decision rules. This may lead to bad out-of-sample predictive performance. High variance is the main reasons that simple regression trees are generally not used. Instead, ensemble methods are preferred.

\section{Frequentist ensemble methods}
\label{frequentist}

We highlighted that individual decision trees tend to overfit and to be sensitive to the training data they were built on. To counteract this, ensemble methods were proposed where multiple models
are aggregated or averaged over to give a more stable and
generalisable solution.

In general, ensemble methods combine the fit from many (hundreds, thousands) simpler models to get an overall predictor. 
All the models we are going to discuss, namely Bagging, Boosting, Random Forest and Bayesian Additive Regression Trees are ensemble methods, for which the simpler models consist of regression trees.

\subsection{Bagging}
Bagging, which stands for Bootstrap Aggregation, is a general technique that aims at reducing the variance of a statistical learning method. It is baseed on the fact that, given independent observation, averaging over them reduces variance.

For the case of regression trees, the main problem of using a single tree as a prediction method is that perturbing the learning set can cause significant changes in the predictor constructed. 
We would thus like to be able to take many different training sets from the population, to build a different trees for each training set and then to average the resulting predictions. However this is unfeasible since we do not have access to many training sets.

Hence, we resort to bootstrap. Bootstrapping consists of taking repeated samples, with replacement, from the single available training set. In such a way, we are able to construct $B$ different bootstrapped training sets. Bagging involves creating multiple $B$ copies of the original training set using hte bootstrap, fitting a separate decision tree to each copy. Then it combines the results from each of the $B$ trees to get an overall prediction.
The single predictions for the target function $f(X)$ for each sample are averaged, as
\begin{equation}
  \label{eq_bagging}
  \hat{f}_{\text{bag}}(x) = \frac{1}{B} \sum_{b=1}^{B} \hat{f}^{*b}(x).
\end{equation}
where $B$ is the number of samples of the bootstrapping procedure.

In bagging the trees are grown big and are not pruned. Therefore, each single tree has high variance and low bias. Averaging these different trees reduce the variance.  Bagging can dramatically reduce the variance of unstable procedures such
as trees, leading to improved prediction.

The drawback of bagging is that we lose interpretability to improve accuracy. When we bag a large number of trees, we can no longer represent the model with a single tree.

\subsection{Random Forest}
A possible problem with bagging is due to the lack of randomization between trees. Indeed, since trees are separately fitted on the bootstrapped training set, it may happen that they share to great amount the same decision rules. This is the case, for example, if there is very strong predictor in the dataset. If this happens, in the collection of bagged trees a vast majority of the trees will use this strong predictor in the top split and they will all be hihgly correlated among themselves.

Correlation among the trees decrease the effectiveness of averaging, as it does not lead to a large reduction in variance, making the entire bagging algorithm less useful.

 Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees. Rather than searching over all the covariate space for each decision rule, as the greedy building of the big trees imposes, in random forests each time we make a split in a tree we randomly sample a subset of $m$ covariates to search over. Therefore, the split is allowed to consider only one of those $m$ predictors. Moreover, at each split is considered a new sample of $m$ predictors. This makes the big trees explore more the space of models.

 Usually, we choose $m \approx \sqrt{p}$.
On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have more of a chance. Therefore, we can think of this process as decorrelating the trees, making the average of the resulting trees less variable and hence more reliable.

@@

Differently from bagging and random forests, boosting does not involve fitting independent trees from many bootstrap samples and averaging the predictions to reduce variance. Instead, it relies on fitting trees sequentially, that is, using information from previously grown trees. Indeed, trees are not fit to bootstrapped versions of the original datasets, but rather to the residual not explained by the remainder of
the trees. Then, its prediction is added, with a weight, to the prediction of the previous trees. This procedure is iterarated until we express the overall prediction of the boosted trees as trhe weighted average of the prediction of the single trees.
This is shown in algorithm \ref{alg_boosting}. 
\begin{algorithm} [h]
  \caption{Boosting for Regression Trees}
  \label{alg_boosting}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}

  Set $\hat{f}(x) = 0$ and $r_i = y_i$ for all $i$ in the training set.\;
  \For{$b = 1, 2, \ldots, B$}{
      (a) Fit a tree $f^b$ with $d$ splits ($d+1$ terminal nodes) to the training data $(X, r)$.\;
      (b) Update $\hat{f}$ by adding in a shrunken version of the new tree:
      \begin{equation}
      \hat{f}(x) \gets \hat{f}(x) + \lambda f^b(x).
      \end{equation}\;
      (c) Update the residuals:
      \begin{equation}
      r_i \gets r_i - \lambda f^b(x_i).
      \end{equation}\;
  }
  Output the boosted model:
  \begin{equation}
  \hat{f}(x) = \sum_{b=1}^B \lambda f^b(x).
  \end{equation}\;

@@

Since each tree \( g(x; T_j, M_j) \) in \eqref{likelihood}, including the variables to split on, the splitting values, and the mean parameters at each terminal node, is unknown, a prior distribution 
\[
p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma)
\]
over all the parameters of the sum-of-trees model is needed.

>>>>

\usepackage{listings}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{listing}
\usepackage{color}
\setlength{\parindent}{0pt}
\setlength{\parskip}{1em}
\usepackage{booktabs}
\usepackage{caption}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{subcaption}
\usepackage{bookmark}
\usepackage[
    backend=biber,
    style=authoryear,
  ]{biblatex}

@@

\title{BART: A Comparison with Frequentist Tree Methods}
\author{
    Vincenzo Dorrello 
    \and
    Giulio Frey 
    \and
    Guido Rossetti 
    \and
    Giovanni Scarpato 
}
\date{\today}

@@

\begin{abstract}
This paper provides a comparison between Bayesian Additive Regression Trees (BART) and traditional non-Bayesian tree-based methods for regression problems. We begin by examining the theoretical foundations of decision trees. We then turn to frequentist ensemble methods, including bagging, random forests, and boosting. Finally, we introduce BART, a Bayesian nonparametric approach that combines the flexibility of regression trees with the formal uncertainty quantification of Bayesian inference. The paper details BART's likelihood, regularization prior, and the Bayesian backfitting MCMC algorithm used for posterior inference. Our final section is a data application, in which we compare the predictive performance of BART against other ensemble methods through both simulated and real-world data. In particular, using the UCI Abalone dataset, we show that BART achieves superior prediction accuracy compared to random forests, boosting, and bagging when using sufficient numbers of trees, though at a higher computational cost. Overall, our findings suggest that BART's automatic prior-based regularization and ability to quantify uncertainty is particularly well suited for complex non-linear modeling tasks.
\end{abstract}

@@

The challenge of modeling complex, non-linear relationships in data has led to the development of various tree-based methods in statistical learning. While traditional parametric approaches like linear regression impose strict assumptions on the relationship between predictors and outcomes, tree-based methods offer greater flexibility in capturing non-linear patterns and interaction effects. This paper focuses on comparing Bayesian Additive Regression Trees (BART) with traditional non-Bayesian tree methods, examining both their theoretical foundations and practical performance.

Tree-based methods have evolved from simple decision trees to ensemble approaches such as bagging, random forests, and boosting. These methods combine multiple trees to improve prediction accuracy and reduce overfitting. However, most traditional approaches lack formal uncertainty quantification and require careful tuning of regularization parameters, for example boosting. BART, introduced by \cite{chipmanBARTBayesianAdditive2010}, addresses these limitations by providing a Bayesian framework that automatically handles regularization through prior specifications while offering uncertainty estimates.

@@

The paper is organized as follows: Section \ref{decision} introduces decision trees and their fundamental concepts. Section \ref{frequentist} explores frequentist ensemble methods including bagging, random forests, and boosting. Section \ref{bart} provides a detailed examination of BART, including its model specification, prior structure, and posterior sampling approach. Section \ref{data_app} presents empirical results from both simulated and real-world data applications. Finally, Section \ref{conclusion} concludes comparing all the tree ensemble methods both on the theoretical and empirical ground. \footnote{The code repository for this paper is available at this \href{https://github.com/giuliofrey/bart}{link}}

@@

Consider a target variable \( y \) and a \( p \)-dimensional vector of covariates, \( \mathbf{x} = (x_1, \ldots, x_p) \). Decision trees are predictive models that make predictions about \( y \) by partitioning the predictor space—the set of all possible values of the random vector \( \mathbf{x} \)—into simple, non-overlapping regions. Notice that also linear
regression in a sense partitions the predictor space, but restricts divisions to certain planes. To the contrary, trees only require that the partition can be achieved by successive splits of the predictor space based on the different predictors.

The structure of a decision tree reflects this partitioning process.We consider only binary trees, where each split of the predictor space can create only two regions. This is in general a good strategy, as multiple splits can be achieved by a series of binary splits. Each tree begins with a root node at the top, representing the entire dataset (the full covariate space). The initial split divides the covariate space into two non-overlapping subsets, according to a decision rule based on the predictor variables \( (x_1, \ldots, x_p) \). Each of these two subsets may be in turn further split into two new (sub-)subsets, and so on. Each split is represented by a new node in the tree. Nodes that represents non-finals splits are called \textit{internal nodes}, while the segments of the tree connecting these nodes are called \textit{branches}. This splitting process, known as \textit{recursive partitioning}, proceeds iteratively until a termination condition is met, such as when further splitting no longer improves predictions or the subsets become too small. The final nodes, called \textit{terminal nodes} or \textit{leaves}, correspond to the smallest subsets created by the tree.

The terminal nodes define a partition of the predictor space into \( J \) non-overlapping regions, \( R_1, R_2, \ldots, R_J \). Each region corresponds to a leaf of the tree. To make a prediction for a given covariate vector \( \mathbf{x}_i \), the tree is traversed starting at the root node. At each internal node, the splitting rule determines whether to proceed to the left or right child node. This process continues until a terminal node is reached. The predicted value \( \hat{y}_i \) is then assigned based on the final region \( \mathbf{x}_i \) falls into.

When \( y \) is continuous, the tree is called a \textit{regression tree}, and the predicted value \( \hat{y}_i \) for a new observation \( \mathbf{x}_i \) falling into region \( R_j \) is the mean of \( y \) values for the training observations in \( R_j \). For categorical \( y \), the tree is referred to as a \textit{classification tree}, and the predicted value is the majority class of \( y \) among the training observations in \( R_j \). Collectively, these approaches are known as \textit{CART} (Classification and Regression Trees), a term introduced by \cite{Gordon_Breiman_Friedman_Olshen_Stone_1984}.

In this work, we focus specifically on regression trees, which aim at predicting a continuous variable \( Y \) given a training dataset. Regression trees differ fundamentally from linear regression models because they do not impose any parametric form on the relationship between \( \mathbf{x} \) and \( Y \). Instead, regression trees use a piecewise-constant approximation of this relationship.

To see this, suppose that the training data is composed by $n$ iid samples
\(\mathbf{X} = \{\mathbf{x}_i\}_{i=1}^n\), where \(\mathbf{x} \in \mathbb{R}^p\), 
along with corresponding response variable
\(Y = \{y_i\}_{i=1}^n\), where \(y_i \in \mathbb{R}\). The relationship between the covariates and the response variable can be expressed as:
\begin{equation}
y = f(\mathbf{x}) + \epsilon \label{regression}
\end{equation}
where \( \epsilon \) is an error, whose distribution is not necessarily specified. Both regression trees and linear regression models seek to estimate the unknown function \( f(\mathbf{x}) \) using the observed data.

The manner in which \( f(\mathbf{x}) \) is approximated depends on the model chosen. Linear regression assumes a parametric form:
\[
f(\mathbf{x}) = \beta_0 + \sum_{j=1}^p \beta_jx_j ,
\]
where \( \beta_0, \beta_1, \ldots, \beta_p \) are coefficients estimated from the data. In contrast, regression trees approximate \( f(\mathbf{x}) \) as a stepwise constant function:
\[
f(\mathbf{x}) = \sum_{j=1}^J \mu_j \cdot 1(\mathbf{x}\in R_j),
\]
where \( R_1, \ldots, R_J \) are the regions defined by the leaves of the tree, and \( \mu_j \) is the mean of \( y \) for the training observations in region \( R_j \).

The choice between these models depends on the nature of the underlying relationship between the predictor variables and the response variable. If the relationship is approximately linear, a linear regression model is likely to outperform a regression tree. Conversely, if the relationship is highly nonlinear, regression trees can capture such patterns more effectively.

\subsection{Regression Trees}

This section draws from \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021} and \cite{Genuer_Poggi_2020}.

A regression tree splits the predictor space into $J$ non-overlapping regions $R_j$. Given a new observation $\mathbf{x}$ that falls in a region $R_j$, the model predicts the corresponding outcome variable $\hat{y}$ to be the mean of the outcome variables corresponding to the observations in the training set that fall in that same region.
 We denote by \( T \) the (binary) tree structure, consisting of a set of interior node decision rules and a set of $b$ terminal nodes. The decision rules are binary splits of the predictor space of the form \( \{\mathbf{x} \in A\} \) vs \( \{\mathbf{x} \notin A\} \), where \( A \) is a subset of the predictor space. Following  \cite{chipmanBARTBayesianAdditive2010}, we only consider rules based on the single components of \( \mathbf{x} = (x_1, \dots, x_p) \) and of the form \( \{x_i \leq c\} \) vs \( \{x_i > c\} \) for continuous \( x_i \). Let \( M = \{\mu_1, \mu_2, \dots, \mu_b\} \) denote the set of parameter values associated with each of the \( b \) terminal nodes of \( T \).

The first step to build a regression tree is to construct the partition $R_1,..., R_j$. Typically, regions are chosen as $p$-dimensional boxes, where $p$ is the number of regressors. The partition is chosen is such a way to minimize the sum of squared errors:  
\begin{equation}
  \{R_1,...,R_J \}=\operatorname*{argmin}_{ \{R_j\}_{j=1}^J}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\mu_{R_j}\right)^2
  \label{rss_tree}
\end{equation}
where $\mu_{R_j}$ is the mean of the training observation in the $j$-th box.

It is computationally infeasible to do the optimization problem in \eqref{rss_tree}. Instead of considering all the possible partitions of the predictor space, a top-down, greedy algorithm, known as recursive binary splitting, is generally used. This approach is \textit{top-down} because it begins at the top of the tree, and then sequentially splits the predictor space. It is also \textit{greedy}, since for each step we choose the split that generates the greatest immediate reduction of the RSS: we do not choose the best sequence of steps looking ahead. To perform the splitting, the algorithm considers the set of all predictors $ \mathbf{x}= (x_1, \ldots, x_p)$, and the set of all possible values of the cutpoint $s$, and then choose the predictor $x_j$ and the cutpoint $s$ such that the splitting generated by the decision rule  $$\{\mathbf{x} \mid x_j < s\} \quad \text{and} \quad \{\mathbf{x} \mid x_j \geq s\}$$ minimizes the RSS.
The process of looking for the predictor and the cutpoint such that the implied decision rule minimize the RSS continues within each subregion of the space until a stopping condition is reached; the more common requires a minimum number of observation within each splitting nodes.

\subsection{Pros and cons of decision trees}
Overall, regression trees offer much greater flexibility in handling the relation between covariates and output than linear models do. This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns and interaction effects between covariates. Greater flexibility comes in handy also when dealing with missing values. Tree offer an effective way to overcome this issue, by constructing surrogate variables.
Moreover, trees are visually easy to interpret. Thanks to their structure, is straightforward to understand why, for a given $x$, we expect a particular value of $y$, following a series of decision rules. Moreover, trees build in such a way are a computationally fast method, that easily scales with large datasets.

However, one of the main disadvantages of single trees is that trees are non-robust. In general, they suffer from high variance: a small change in the data can lead to a large change in the estimation. The main reason of this instability is the hierarchical structure of the process as the effect of an error in one of the first nodes is propagated to the other nodes below it. Indeed, recursive binary splitting is likely to produce good prediction on the training-set, but overfit the training data, leading to poor out-of-sample predictive performance. The reason is that this procedure is likely to generate a complex tree, with a low bias but high variance in prediction. A possible solution could be constraining ex ante the size of trees, generating a small tree with low variance but possibly high bias, since potentially good splits further down in tree would possibly remain unexploited. A better strategy, called \textit{weakest link pruning}, is to grow a very large tree $T_0$, and then prune it back in order to obtain a subtree.

However, the predictive performance of pruned trees still falls short of what models combining multiple trees can achieve. These models are called tree ensemble methods and are discussed in the following section.

\section{Frequentist tree ensemble methods}
\label{frequentist}

We highlighted that individual decision trees tend to overfit and their prediction to be highly sensitive to the training data they were built on. A general solution, not only in the case of trees, to improve predictive performance of a model are ensemble methods. Ensemble methods combine the fit from many (hundreds or thousands) simpler models to get an overall predictor. 
All the models we are going to discuss, namely Bagging, Boosting, Random Forest and Bayesian Additive Regression Trees are ensemble methods, for which the simpler models consist of regression trees.

\subsection{Bagging}
Bagging, which stands for Bootstrap Aggregation, is an ensemble method that aims at reducing the variance of a generic statistical model. It is based on the fact that, given independent observation, averaging over them reduces variance.
For the case of regression trees, the main problem of using a single tree as a prediction method is that perturbing the learning set can cause significant changes in the predictor constructed. We would thus like to be able to take many different training sets from the population, to build a different trees for each training set and then to average the resulting predictions. However this is unfeasible since we do not have access to many training sets.

Bootstrapping consists of taking repeated samples, with replacement, from the single available training set to simulate sampling many different training sets. In such a way, we are able to construct $B$ different bootstrapped training sets. In bagging we fit a separate decision tree to each copy of the original training set, and then average the predictions $\hat{y}$ from each of the $B$ trees to get an overall prediction.
The single predictions for the target function $f(\mathbf{x})$ for each sample are averaged, as
\begin{equation}
  \label{eq_bagging}
  \hat{f}_{\text{bag}}(\mathbf{x}) = \frac{1}{B} \sum_{b=1}^{B} \hat{f}^{*b}(\mathbf{x}).
\end{equation}
where $B$ is the number of samples of the bootstrapping procedure.

In bagging the trees are grown big and are not pruned. Therefore, each single tree has high variance and low bias. Averaging these different trees reduce the variance. However, the drawback of bagging (and of ensemble methods in general) is the loss of interpretability required in order to improve accuracy. Indeed, when we bag a large number of trees, we can no longer represent the model with a single tree and directly interpret the relations between covariates throught the structure of this tree.

\subsection{Random Forest}
A possible problem with bagging is due to the lack of randomization between trees. Indeed, since trees are separately fitted on the bootstrapped training set, it may happen that they share to great amount the same decision rules. This is the case, for example, if there is very strong predictor in the dataset. If this happens, in the collection of bagged trees a vast majority of the trees will use this strong predictor in the top split and they will all be highly correlated among themselves. Correlation among the trees decrease the effectiveness of averaging, as it does not lead to a large reduction in variance as it would be the case with independent trees. This makes the entire bagging algorithm less useful.

Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees. Rather than searching over all the covariate space for each decision rule, as the greedy algorithm used for building of the single trees imposes, in random forests each time we make a split in a tree we randomly sample a subset of $m$ covariates to search over. Therefore, the split is allowed to consider only one of those $m$ predictors. Moreover, at each split is considered a new sample of $m$ predictors.  On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have a higher probability to be selected. This increases the exploration of the  space of models and reduces correlation among trees, making the average of the resulting trees less variable and hence more reliable.

@@

Differently from bagging and random forests, boosting does not involve fitting independent trees from many bootstrap samples and averaging the predictions to reduce variance. Instead, it relies on fitting trees sequentially, that is using information from previously grown trees. Indeed, trees are not fit to bootstrapped versions of the original datasets, but rather to the residual not explained by the remainder of the trees. Then, each new tree's prediction is added, with a weight, to the prediction of the previous trees. This procedure is iterated until the prediction of the boosted trees is the weighted average of the prediction of all the single trees.
This is shown in algorithm \ref{alg_boosting}, adapted from \cite{jamesIntroductionStatisticalLearning2021}.
\begin{algorithm} [h]
  \caption{Boosting for Regression Trees}
  \label{alg_boosting}
  \SetAlgoLined
  \DontPrintSemicolon
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}

  Set $\hat{f}(\mathbf{x}) = 0$ and $r_i = y_i$ for all $i$ in the training set ($i= 1, \ldots, n$).\;
  \For{$b = 1, 2, \ldots, B$}{
      (a) Fit a tree $f^b$ with $d$ splits ($d+1$ terminal nodes) to the training data $[(\mathbf{x}_1, \ldots \mathbf{x_n)}, (r_1, \ldots, r_n)]$.\;
      (b) Update $\hat{f}$ by adding in a shrunken version of the new tree:
      \begin{equation}
      \hat{f}(\mathbf{x}) \gets \hat{f}(\mathbf{x}) + \lambda f^b(\mathbf{x}).
      \end{equation}\;
      (c) Update the residuals:
      \begin{equation}
      r_i \gets r_i - \lambda f^b(\mathbf{x}_i).
      \end{equation}\;
  }
  Output the boosted model:
  \begin{equation}
  \hat{f}(\mathbf{x}) = \sum_{b=1}^B \lambda f^b(\mathbf{x}).
  \end{equation}\;

@@

Since each tree \( g(\mathbf{x}; T_j, M_j) \) in \eqref{likelihood}, including the variables to split on, the splitting values, and the mean parameters at each terminal node, is unknown, a prior distribution 
\[
p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma)
\]
over all the parameters of the sum-of-trees model is needed.
